---
output: github_document
bibliography: vignettes/REFERENCES.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/"
)
```


# SPLITT: Serial and Parallel LIneage Traversal of Trees

SPLITT is a one-header C++ library that makes it easy to program fast pre-order and post-order traversal algorithms over tree-like data. The library is written in C++11. If [OpenMP](https://www.openmp.org) is enabled during compilation, SPLITT can take advantage of a multi-core CPU and/or vectorized instructions to execute the tree traversal in parallel. 

## The SPLITT.h file

The SPLITT library represents a C++ header file called "SPLITT.h". Technically, downloading [the latest version of the file from github](https://github.com/venelin/SPLITT/raw/master/src/SPLITT.h) and including it in one or more of your C++ files is all you need to start using it. However, before you can do anything useful with SPLITT you will need to get familiar with some of its classes, i.e. its programming interface. 

## The SPLITT R-package

The R-package called with the same name (SPLITT) bundles the SPLITT.h header together with its documentation and examples. You don't need to install this package nor do you need to know R or to have an R environment on your system for using the SPLITT library. 

## User guides and resources

The user guides and the technical reference for the library are available from the [SPLITT web-page](https://venelin.github.io/SPLITT/index.html). 

The research article "Parallel Likelihood Calculation for Phylogenetic Comparative Models: the SPLITT C++ Library" provides a general overview of SPLITT. The example application to the PMM model depicted on Fig. 1 in this article is used as a coding example in the SPLITT online documentation. The article is currently in peer review for a publication and is available as a preprint from [BioRxiv](https://www.biorxiv.org/content/early/2018/10/29/235739).

The SPLITT source code is located in the [SPLITT github repository](https://github.com/venelin/SPLITT). 

Feature requests, bugs, etc can be reported in the [SPLITT issues list](https://github.com/venelin/SPLITT/issues).

## Citing SPLITT

To give credit to the SPLITT library in a publication, please cite the following article:

Mitov, V., & Stadler, T. (2018). Parallel Likelihood Calculation for Phylogenetic Comparative Models: the SPLITT C++ Library. bioRxiv, 235739. http://doi.org/10.1101/235739

## Used R-packages
```{r create-references, echo=FALSE, include=FALSE, eval=TRUE}
treeProcessing <- c("ape")
testing <- c("testthat")
Rcpp <- c("Rcpp")
RcppArmadillo <- c("RcppArmadillo")

packagesUsed <- c(treeProcessing, Rcpp, RcppArmadillo, testing)

printPackages <- function(packs) {
  res <- ""
  for(i in 1:length(packs)) {
    res <- paste0(res, paste0(packs[i], ' v', packageVersion(packs[i]), ' [@R-', packs[i], ']'))
    if(i < length(packs)) {
      res <- paste0(res, ', ')
    }
  }
  res
}

# Write bib information (this line is executed manually and the bib-file is edited manually after that)
knitr::write_bib(packagesUsed, file = "./vignettes/REFERENCES-R.bib")
```

The SPLITT R-package uses the following 3rd party R-packages:

* For calling C++ from R: `r printPackages(c(Rcpp, RcppArmadillo))`
* For tree processing in R: `r printPackages(treeProcessing)`;
* For unit-testing: `r printPackages(testing)`.

## References

