---
title: "Getting started with the SPLITT library"
author: "Venelin Mitov"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: REFERENCES.bib
---

```{r setup, include = FALSE}
# Make results reproducible
set.seed(1)
knitr::opts_chunk$set(cache = FALSE)
options(digits = 4, width=120, scipen = 999)
remove.packages("SPLITT")
```

# Before you start
Before you consider using the **SPLITT** C++ library, try to answer the following 
questions for yourself.

## Do you use pre-order or post-order traversal of a tree-like data-structure?
If you don't understand what I mean, chances are high that you simply do not 
need to deal with **SPLITT** at that point. The reason is that I wrote 
**SPLITT** to solve a very specific problem, which I've encountered many times
during my doctoral studies: perform the same type of calculation for all nodes 
in a *large tree*, where each node is related to some specific data and the 
calculation for one node uses as input the result from the calculation performed
on its *daughter nodes*. This is what I mean by *post-order traversal*. A 
*pre-order traversal* is the same with only one difference -- replace the term
*daughter nodes* with the term *parent node* in the sentence above. 

## How big are your trees and how complex is your node traversal operation?
When I said *large tree* I meant a tree of more than 100 tips, sometimes, more 
than 10000 tips. This is not a hard rule of thumb, but typically parallel 
traversal has a performance benefit only on big trees. This depends on other 
factors as well, such as the computational and memory complexity of the
node traversal operation. If the tree is relatively small and the operation is
very simple (e.g. the addition or multiplication of a few double-precision 
numbers), then it is very likely that using **SPLITT** in parallel mode would 
not be faster than using it in serial mode. Yet, using **SPLITT** might still 
be a good idea if you currently have an implementation in R and wish to have a
faster C++ version - in this case a serial **SPLITT**-based implementation is 
probably going to between 10 and 100 times faster than the R-implementation. 

## Do you need the tree-traversal to be fast?
This very much depends on how frequently do you need to perform the 
tree-traversal. If you only need to do it once or a couple of times with a given 
dataset, then I don't see much need of speeding this up. In my specific use-case,
I needed to run a tree traversal millions of times and this was by far the
heaviest computational operationduring the analysis of my biological data. It  
would not be an exaggeration, if I say that,without **SPLITT**, my PhD would
have needed several more decades of computation time on a modern high performance 
computing cluster. Thanks to the fast C++ implementation of my models, based on 
**SPLITT**, I was able to tune my models by running many more analyses with 
different parameter settings, such as prior distributions for a Bayesian model
inference. 

## Do you have a program solving your computational problem in a higher level language, e.g. R?
If you don't, I recommend not trying **SPLITT**, before you have written that 
program. One reason for that is that, as every C++ code, **SPLITT** needs to 
be compiled before it can be run. Because you will be including the file "SPLITT.h"
in your program, it will be re-compiled every time you make a change in your program
and wish to test it. Typically, this would needs about two minutes. Hence, you 
risk to spend hours or days if you are heavily debugging your C++ code. Of course,
if you only know how to program in C++, then no need to learn R now - stick to 
C++ but be prepared with some snacks and coffee. 

# The easiest way to start using **SPLITT**
Being a software library **SPLITT** does not provide a high-level end-user interface. 
Rather, **SPLITT** is used via its application programming interface (API). 
In other words, you use the library by putting your application specific code in 
several C++ classes. The easiest way to do this is to start from an example of 
such classes and use this example as a skeleton for your own code, i.e. replace 
the code in the example with your application specific data-types, and traversal
operations. Because I was using **SPLITT** mostly from an R-package, at present, 
there is only an R-package example. Even if you are not familiar with R, this 
guide should still help you understand how to use **SPLITT** from within your
C++ program. 

## Installing the **SPLITT** R-package
I recommend installing the latest package version from github. From the R-prompt,
type the following command:

```{r, echo=TRUE, eval=FALSE}
devtools::install_github("venelin/SPLITT")
```


```{r, echo=FALSE, eval=TRUE}
capture <- capture.output( devtools::install_github("venelin/SPLITT") )
capture <- 'Building SPLITT vignettes
* installing *source* package ‘SPLITT’ ...
** libs
/usr/local/clang6/bin/clang++  -I"/Library/Frameworks/R.framework/Resources/include" -DNDEBUG -fopenmp -I"/Users/vmitov/Library/R/3.5/library/Rcpp/include" -I"/Users/vmitov/Library/R/3.5/library/RcppArmadillo/include" -I/usr/local/include   -fPIC  -Wall -g -O2  -c RCPP__AbcPMM.cpp -o RCPP__AbcPMM.o
/usr/local/clang6/bin/clang++  -I"/Library/Frameworks/R.framework/Resources/include" -DNDEBUG -fopenmp -I"/Users/vmitov/Library/R/3.5/library/Rcpp/include" -I"/Users/vmitov/Library/R/3.5/library/RcppArmadillo/include" -I/usr/local/include   -fPIC  -Wall -g -O2  -c RCPP__Tree.cpp -o RCPP__Tree.o
/usr/local/clang6/bin/clang++  -I"/Library/Frameworks/R.framework/Resources/include" -DNDEBUG -fopenmp -I"/Users/vmitov/Library/R/3.5/library/Rcpp/include" -I"/Users/vmitov/Library/R/3.5/library/RcppArmadillo/include" -I/usr/local/include   -fPIC  -Wall -g -O2  -c RcppExports.cpp -o RcppExports.o
/usr/local/clang6/bin/clang++ -dynamiclib -Wl,-headerpad_max_install_names -undefined dynamic_lookup -single_module -multiply_defined suppress -L/Library/Frameworks/R.framework/Resources/lib -L/usr/local/lib -o SPLITT.so RCPP__AbcPMM.o RCPP__Tree.o RcppExports.o -fopenmp -F/Library/Frameworks/R.framework/.. -framework R -Wl,-framework -Wl,CoreFoundation
installing to /Users/vmitov/Library/R/3.5/library/SPLITT/libs
** R
** tests
** byte-compile and prepare package for lazy loading
** help
*** installing help indices
*** copying figures
** building package indices
** testing if installed package can be loaded
* DONE (SPLITT)'

cat(capture)
```

## See if it works on your computer
```{r}
library(SPLITT)
MiniBenchmark(N=10, Ntests = 1000)
```





```{r treeAndDataLogLik}
library(ape)
newick <- '(((1:3,2:2.8)6:4.2,3:4.1)7:4.5,(4:4,5:5)8:6)0;'
tree <- read.tree(text=newick)

x0 <- 0.1
sigma2 <- 0.25
sigmae2 <- 1

set.seed(1)
g <- rTraitCont(tree, model = "BM", root.value = x0, 
                sigma = sqrt(sigma2),
                ancestor = FALSE)

x <- g + rnorm(n = length(tree$tip.label), mean = 0, sd = sqrt(sigmae2))

PMMLogLikCpp(x, tree, x0, sigma2, sigmae2, mode = 1)
```

```{Rcpp, eval=FALSE, echo=TRUE}
#ifndef AbcPMM_H_
#define AbcPMM_H_

#include "./SPLITT.h"
#include "./NumericTraitData.h"
#include <iostream>
#include <cmath>

namespace SPLITT {

template<class Tree>
class AbcPMM: public TraversalSpecification<Tree> {

public:
  typedef AbcPMM<Tree> MyType;
  typedef TraversalSpecification<Tree> BaseType;
  typedef Tree TreeType;
  typedef PostOrderTraversal<MyType> AlgorithmType;
  typedef vec ParameterType;
  typedef NumericTraitData<typename TreeType::NodeType> DataType;
  typedef vec StateType;

  double sigmae2, sigma2;
  vec x;
  vec a, b, c;

  AbcPMM(TreeType const& tree, DataType const& input_data):
    BaseType(tree) {

    if(input_data.x_.size() != this->ref_tree_.num_tips()) {
      std::ostringstream oss;
      oss<<"The vector x must be the same length as the number of tips ("<<
        this->ref_tree_.num_tips()<<"), but were"<<input_data.x_.size()<<".";
      throw std::invalid_argument(oss.str());
    } else {

      uvec ordNodes = this->ref_tree_.OrderNodes(input_data.names_);
      this->x = At(input_data.x_, ordNodes);
      this->a = vec(this->ref_tree_.num_nodes());
      this->b = vec(this->ref_tree_.num_nodes());
      this->c = vec(this->ref_tree_.num_nodes());
    }
  };

  StateType StateAtRoot() const {
    vec res(3);
    res[0] = a[this->ref_tree_.num_nodes() - 1];
    res[1] = b[this->ref_tree_.num_nodes() - 1];
    res[2] = c[this->ref_tree_.num_nodes() - 1];
    return res;
  };

  void SetParameter(ParameterType const& par) {
    if(par.size() != 2) {
      throw std::invalid_argument(
      "The par vector should be of length 2 with \
      elements corresponding to sigma2 and sigmae2.");
    }
    if(par[0] <= 0 || par[1] <= 0) {
      throw std::logic_error("The parameters sigma2 and sigmae2 should be positive.");
    }
    this->sigma2 = par[0];
    this->sigmae2 = par[1];
  }

  inline void InitNode(uint i) {
    
    if(i < this->ref_tree_.num_tips()) {
      a[i] = -0.5 / sigmae2;  
      b[i] = x[i] / sigmae2;
      c[i] = -0.5 * (x[i]*b[i] + log(2*G_PI*sigmae2));
    } else {
      a[i] = b[i] = c[i] = 0;
    }
  }

  inline void VisitNode(uint i) {
    double t = this->ref_tree_.LengthOfBranch(i);
    
    double d = 1 - 2*a[i]*sigma2*t;
    // the order is important here because for c[i] we use the previous values 
    // of a[i] and b[i].
    c[i] = c[i] - 0.5*log(d) + 0.5*b[i]*b[i]*sigma2*t/d;
    a[i] /= d;
    b[i] /= d;
  }

  inline void PruneNode(uint i, uint j) {
    a[j] = a[j] + a[i];
    b[j] = b[j] + b[i];
    c[j] = c[j] + c[i];
  }

};

typedef TraversalTask<
  AbcPMM<OrderedTree<uint, double>> > ParallelPruningAbcPMM;
}
#endif //AbcPMM_H_
```

```{r fig1a, dpi=150, fig.width=4.6}
par(mfrow=c(1,5))
par(mar=c(0,0,0,0))
par(oma=c(0, 0,0, 0))

PlotParallelTraversal(tree)
```


